---
title: "Phyloseq"
output: pdf_document
date: "2024-04-18"
---

#Load required packages
```{r}
library(phyloseq)
library(BiMiCo)
library(dplyr)
library(devtools)
library(MicEco)
library(ggplot2)
library(vegan)
```

# Load taxa and seqtab.nochim
```{r}
load("RData/taxa")
load("RData/seqtab.nochim.RData")
```

# Impot metatata
```{r}
metadata<-read.csv("sample-metadata.csv", header=TRUE, row.names = 1)
```


# create phyloseq object
```{r}
#load data first
phyloseq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE), 
                     sample_data(metadata),
                     tax_table(taxa))
#show file
phyloseq
```

# transform sample counts
```{r}
#convert from row to abundance
phyloseq <- transform_sample_counts(phyloseq, function(abund) 1*(abund>0))
phyloseq
```

# remove the sequence itself and replace with ASV
```{r}
dna <- Biostrings:: DNAStringSet(taxa_names(phyloseq))
names(dna) <- taxa_names(phyloseq)
phyloseq <- merge_phyloseq(phyloseq, dna)
taxa_names(phyloseq) <- paste0("ASV", seq(ntaxa(phyloseq)))
phyloseq
```

# remove mitochondria and chloroplast matches. remove all non bacterial sequences
```{r}
physeq <- phyloseq %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) &
                                  Order!= "Chloroplast" | is.na(Order) )
physeq
```

# Remove all non bactrial sequences
```{r}
physeq<-rm_nonbac(physeq)
physeq
```

# Save physeq objects to start
```{r}
save(physeq, file="RData/physeq.RData")
```

# load physeq objects
```{r}
load("RData/physeq.RData")
```

# Plot bar graph based on phylum
```{r}
plot_bar(physeq, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat= "identity", position= "fill")
```

# Create a barplot of relative abundance
```{r}
# convert to relative abundance
physeq_relabund <- transform_sample_counts(physeq, function(x) x / sum(x))

# bar plot
plot_bar(physeq_relabund, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat= "Identity", position= "fill") + facet_wrap(~body.site, scales = "free")
```

# plot alpha diversity based on body site
```{r}
plot_richness(physeq, x="body.site", color= "subject", measures=c("Observed", "Simpson", "Shannon"))
```

# pLot alpha diversity based on antibiotic usage
```{r}
plot_richness(physeq, x="reported.antibiotic.usage", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
```

# Plot alpha diversity based on subject
```{r}
plot_richness(physeq, x="subject", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
```
# Test for normality
```{r}
 alpha <- estimate_richness(physeq, measures = c ("Observed", "Simpson", "Shannon"))
#creates an object called alpha using estimate richness feature of phyloseq

# Shapiro-wilk
#making three objects: observed, shannon, simpson
observed <- shapiro.test(alpha$Observed)
shannon <- shapiro.test(alpha$Shannon)
simpson <- shapiro.test(alpha$Simpson)

#print
print(observed)
print(shannon)
print(simpson)
```

# create data frames for statistical analysis 
```{r}
# extract sample information from the physeq oject
samples <- sample_data(physeq)

#If samples is a phyloseq sample_data object, convert it to a data frame
if (class(samples) == "sample_data") {
  samples <- data.frame(sample_data(samples)) 
}

# add a column to alpha with sample names 
alpha$sample <- rownames(alpha)

# merge alpha diversity data and sample data 
alpha <- merge(alpha, samples, by = "sample")
```

# perform statistics based on subject
```{r}
 # perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ subject, data = alpha)
test_simpson <- wilcox.test(Simpson ~ subject, data = alpha)
test_shannon <- t.test(Shannon ~ subject, data = alpha)
#use t test for shannon because it was evenly distributed

# Print the results
print(test_observed)
print(test_simpson)
print(test_shannon)
```

# perform statistics based on reported.antibiotic.usage
```{r}
 # perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ reported.antibiotic.usage, data = alpha)
test_simpson <- wilcox.test(Simpson ~ reported.antibiotic.usage, data = alpha)
test_shannon <- t.test(Shannon ~ reported.antibiotic.usage, data = alpha)
#use t test for shannon bc it was evenly distributed

# Print the results
print(test_observed)
print(test_simpson)
print(test_shannon)
```

# test for body site
```{r}
# 3 or more groups need to do anova test for normally distributed data, if not normally distributed do kruskal wallace test

kruskal.test(Simpson ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Simpson, alpha$body.site, p.adjust.method="holm")
kruskal.test(Observed ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Observed, alpha$body.site, p.adjust.method="holm")
shannonanova <- aov(Shannon ~ body.site, data=alpha)
summary(shannonanova)
TukeyHSD(shannonanova)

```

# Remove taxa with relative abundance <0.005%
```{r}
minToRelAbun = .00005
x = taxa_sums(physeq)
keepTaxa = (x/sum(x)) > minToRelAbun
physeqprune = prune_taxa(keepTaxa, physeq)
physeqprune
```

# Number of shared ASVs body.site (found in 25% or more)
```{r}
bodysite=ps_venn(
  physeqprune,
  "body.site",
  fraction = .25,
  weight = FALSE,
  #weighting rare taxa
  relative = TRUE,
  #relative abundance
  plot = TRUE
  #if true, give vinn diagram, if not gives list
)
bodysite
```

# Bray Curtis Calculation
```{r message=FALSE}
set.seed(777)
dist = phyloseq::distance(physeqprune, method= "bray", weighted=TRUE)
oridination = ordinate(physeqprune, method="NMDS", distance=dist)
```

# Bray Curtis subject plot
```{r}
braysubject=plot_ordination(physeqprune, ordination, color="subject") + 
  theme_classic() +
  theme(strip.background = element_blank()) + 
  stat_ellipse(aes(group=subject))
braysubject
```

# Bray Curtis Body Site plot
```{r}
braybodysite=plot_ordination(physeqprune, ordination, color="body.site") + 
  theme_classic() +
  theme(strip.background = element_blank()) + 
  stat_ellipse(aes(group=body.site))
braysubject
```

# Bray Curtis reported.antibiotic.usage plot
```{r}
brayantibiotic.usage=plot_ordination(physeqprune, ordination, color="reported.antibiotic.usage") + 
  theme_classic() +
  theme(strip.background = element_blank()) + 
  stat_ellipse(aes(group=body.site))
braysubject
```

# Bray Curtis Subject Statistics
```{r}
#pulling data from physeqprune based on subject. Testing whether or not significant difference in communities between subjects
adonis2(dist ~ sample_data(physeqprune)$subject)
```

# Bray Curtis Body Site Statisics
```{r}
adonis2(dist ~ sample_data(physeqprune)$body.site)
#need this code because more than 2 things that we are comparing
ps.disper<-betadisper(dist, sample_data(physeqprune)$body.site)
permutest(ps.disper, pair=TRUE)
#look at pairwise comparisons for p values when to compare different sites
```

# Bray Curtis Reported Antibiotic Usage Statistics
```{r}
#pulling data from physeqprune based on subject. Testing whether or not significant difference in antibiotic usage
adonis2(dist ~ sample_data(physeqprune)$reported.antibiotic.usage)
```